---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

Load the data into R
```{r echo=TRUE}
library(dplyr)
library(lubridate)

data <- read.csv("activity.csv", header = TRUE, stringsAsFactors = FALSE)
str(data)

dataset <- tbl_df(data)

## convert date
dataset <- dataset %>% mutate(date=ymd(date))
summary(dataset)
```

## What is mean total number of steps taken per day?

```{r echo=TRUE}
per_day <-  dataset %>% na.omit %>% group_by(date)

summary <- per_day %>% summarise(sum(steps))
names(summary) <- c("day", "total_steps") 

print(summary) 
```

```{r echo=TRUE}
library(ggplot2)

qplot(total_steps, data = summary, geom = "histogram", bins = 40, 
      main = "The total number of steps taken each day")

mean_total_steps <- mean(summary$total_steps)
print(mean_total_steps)

median_total_steps <- median(summary$total_steps)
print(median_total_steps)
```

Mean total number of steps taken per day: **`r as.integer(mean_total_steps)`**
Median total number of steps taken per day: **`r as.integer(median_total_steps)`**

## What is the average daily activity pattern?

```{r echo=TRUE}
per_interval <- dataset %>% na.omit %>% group_by(interval)

summary2 <- per_interval %>% summarise(mean(steps))
names(summary2) <- c("interval", "mean_steps") 
str(summary2)

plot(summary2$interval, summary2$mean_steps, type = "l" ,
     xlab="5-minute interval", ylab="Average number of steps taken",
     main="Daily activity pattern")


## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
print( summary2 %>% filter(mean_steps == max(mean_steps)) )
```



## Imputing missing values

Number of missing values in *steps* column: `r sum(is.na(dataset$steps))`  
Number of missing values in *date* column: `r sum(is.na(dataset$date))`  
Number of missing values in *interval* column: `r sum(is.na(dataset$interval))`  



```{r echo=TRUE}
## average number of steps taken every day
mean_total_steps <- mean(summary$total_steps)
print(mean_total_steps)
```

**Let's replace missing values in *steps* column with an average number of steps taken every day divided by the number of time periods.**

```{r echo=TRUE}
repl <- as.integer(mean_total_steps / length(unique(dataset$interval)))
fixed_dataset <- mutate(dataset, steps = ifelse(is.na(steps), repl,  steps))
str(fixed_dataset)
```

Number of missing values in *steps* column: `r sum(is.na(fixed_dataset$steps))`  

```{r echo=TRUE}
per_day <- fixed_dataset %>% group_by(date)

summary3 <- per_day %>% summarise(sum(steps))
names(summary3) <- c("day", "total_steps") 

print(table(summary3$total_steps))
qplot(total_steps, data = summary3, geom = "histogram", bins = 40,
      main = "The total number of steps taken each day")

## Calculate and report the mean and median total number of steps taken per day. Do these values differ from the
## estimates from the first part of the assignment?

mean_total_steps <- mean(summary3$total_steps)
print(mean_total_steps)

median_total_steps <- median(summary3$total_steps)
print(median_total_steps)
```

The mean and median total number of steps taken per day don't significantly differ from the estimates from the first part of the assignment.

Imputing missing data doesn't have any significant impact on the estimates of the total daily number of steps.


## Are there differences in activity patterns between weekdays and weekends?

```{r echo=TRUE}
fixed_dataset <- fixed_dataset %>% mutate(day_type = ifelse(wday(date) == 1 | wday(date) == 7, "weekend", "weekday"))
str(fixed_dataset)


per_interval <- fixed_dataset %>% group_by(interval, day_type)

summary4 <- per_interval %>% summarise(mean(steps))
names(summary4) <- c("interval", "day_type", "mean_steps") 
str(summary4)

library(lattice)

xyplot(mean_steps ~ interval | day_type, type = "l" ,
     xlab="5-minute interval", ylab="Average number of steps taken",
     main="Daily activity pattern", data=summary4)

```

